Exploring probability distributions for bivariate temporal granularities

### 2020-02-03 main.Rmd
smart meter application - all captions and headers fixed.

<!-- # how data looks -->
Smart meters provide large quantities of measurements on energy usage for households across Australia. One of the customer trial [@smart-meter] conducted as part of the Smart Grid Smart City (SGSC) project (2010-2014) in Newcastle, New South Wales and some parts of Sydney provides customer wise data on energy consumption in `general_supply_kwh` in every half hour from February 2012 to March 2014. \ref{tab:smart-meter} shows the data structure of the available data. It would be interesting to explore the energy consumption distribution for these customers and gain insights on their energy behavior which are lost either due to aggregation or looking only at coarser temporal units. The idea here is to show how looking at the energy consumption across different cyclic granularities together can help identify different behavioral patterns and identify the extreme and regular households.

<!-- We take a random sample of 50 households to see how we can achieve that.  -->

@@

@@

```{r smart-meter, fig.cap = "Smart meter data structure"}

sm <- 
  read_rds("data/sm_cust50.rds")

smart_meter <- sm %>% as_tsibble() %>% dplyr::select(customer_id, 
                              reading_datetime, 
                              general_supply_kwh, 
                              everything()) %>% head()
smart_meter
# knitr::kable(smart_meter, format = "markdown")
```

@@

```{r search, echo=TRUE}
library(gravitas)
smart_meter %>% search_gran()
#smart_smart_search <- smart_meter %>% search_gran()
#knitr::kable(smart_smart_search, format = "markdown")
```

@@

```{r search_limit1, echo=TRUE}
smart_meter10 %>% 
  search_gran(highest_unit = "month")
# smart_meter_limit1 <- smart_meter10 %>% 
#   search_gran(highest_unit = "month")
# knitr::kable(smart_meter_limit1, format = "markdown")
```

@@

```{r search_gran_limit2, echo=TRUE}
smart_meter10 %>% search_gran(highest_unit = "month",
                              filter_out = c("hhour", "fortnight")
                              )
# smart_meter_limit2 <- smart_meter10 %>% search_gran(highest_unit = "month",
#                               filter_out = c("hhour", "fortnight")
# )
# knitr::kable(smart_meter_limit2, format = "markdown")
```

@@

<!-- From the Gestalt theory with the same data  <$C1$ $C2$, `general_supply_kwh`>, mapping $C1$ on the x-axis vs facet might lead to different insights. -->

```{r harmony, echo=TRUE}
smart_meter10 %>% harmony(
  ugran = "month",
  filter_out = c("hhour", "fortnight")
)
# smart_meter10 %>% gran_advice("wknd_wday", "hour_day")

If the data for all keys are visualized together, it might lead to Simpson's paradox, which occurs when one `customer_id` shows a particular behavior, but this behavior is reversed when all keys are combined together. This is also intuitive because `customer_id`'s with very different occupation or demographics will tend to have very different energy behavior and combining them together will somehow weaken any typical or extreme behavior. For example, Figure \ref{fig:wknd-wday50} shows the quantile plot of energy consumption of a sample of 50 households. Quantiles for weekends and weekdays for these customers look exactly the same. It is not clear if the behaviors of most of these customers do not alter between weekdays and weekends or looking at all customers together balances out the behaviors. Consider another example in Figure \ref{fig:my-hd} showing the area quantile plots of 50 customers. The median is very close to the lower boundaries of all the other bands implying energy consumption for these households are left skewed. The pink band changes significantly in winter months (May - August) implying that much of the behavioral changes occur in the quartiles, that too in peak hours of the day. It is to be noted here that the the level of all quantiles increased too in winter months (increased energy usage in weather conditions), the interesting part is to notice that the relationship between bands other than quartile stayed same across seasons.

@@

```{r wknd-wday50, fig.cap = "Quantile plot of energy consumption of 50 households across different hours of the day faceted by weekday and weekend. The quantiles are not different for weekdays and weekends implying either the behavior balances out among these customers or most of them do not behave differently for weekdays and weekends."}

smart_meter <- 
  read_rds("data/sm_cust50.rds")

smart_meter %>% 
  prob_plot("wknd_wday",
                            "hour_day",
                            response = "general_supply_kwh",
                            plot_type = "quantile",
            symmetric = FALSE, 
            quantile_prob = c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99)) +
  scale_y_sqrt()
```

@@

```{r my-hd, fig.cap="Area quantile plots of energy consumption across hours of the day faceted by months of the year. The black line is the median, whereas the pink band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile. The median and quartile consumption increase in winter months, but extreme behavior represented by higher bands mostly stay the same across seasons. "}

library(gravitas)
library(tsibble)

library(ggplot2)
smart_meter %>%
  prob_plot("month_year", "hour_day",
    response = "general_supply_kwh",
    plot_type = "quantile",
    quantile_prob = c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99),
    symmetric = TRUE) +
  ggtitle("") + scale_y_sqrt(breaks = c(0.01, 0.1, 0.5, 1:3))
```

@@

Looking at random samples with different demographics might confound our analysis and a strategy for analyzing multiple keys together could be to first screen some basis time series or demographic features and then look at their energy behavior. This is beyond the scope of the current work. For the purpose of this paper, the smart meter data is filtered for two customers to illustrate kinds of insights that can be drawn after proceeding with the systematic exploration.

@@

In \autoref{fig:bothcust} we visualize the harmony pair (wknd_wday, hour_day) through a box plot. Boxplot of energy consumption is shown across `wknd_wday` (facet) and `hour_day` (x-axis) for the same two households. For the first household, the shape of the distribution does not change significantly for weekdays and weekends. The distribution is roughly smooth with a morning and an evening peak. For the second household, the quantiles behave differently in peak hours. For the first few hours on weekdays, energy consumption starts at a much higher level than weekends. Consider looking at letter value plots instead of quantile plots to look at the same data in \autoref{fig:bothcust-lv}. There is additional information that we can derive looking at the distribution.
There is bimodality in the early hours of the day, implying both low and high energy consumption is probable in the early hours of the day. Also hours from 7 to 13 look most volatile. If we visualize the same data with reverse mapping of the cyclic granularities, then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. For example in \autoref{fig:cust-lv}, it can be seen that median energy consumption for the early morning hours is extremely high for weekdays compared to weekends. Also, outliers are more prominent in the latter part of the day. All of these indicates that looking at different distribution plots or changing the mapping might draw lights on different aspect of the energy behavior for the same customer.

@@

```{r bothcust_notrun}
cust1 <- smart_meter10 %>% 
  filter(customer_id %in% c(10006704)) %>% 
  prob_plot("wknd_wday",
                            "hour_day",
                            response = "general_supply_kwh",
                            plot_type = "quantile",
            symmetric = FALSE,
            quantile_prob = c(0.25, 0.5, 0.75, 0.9)) +
  scale_y_sqrt() +
  ggtitle("Energy consumption distribution for customer id: 10006704")

cust2 <- smart_meter10 %>% 
  filter(customer_id %in% c(10017936))%>% 
  prob_plot("wknd_wday",
                            "hour_day",
                            response = "general_supply_kwh",
                            plot_type = "quantile",
            symmetric = FALSE,
             quantile_prob = c(0.25, 0.5, 0.75, 0.9)) +
  scale_y_sqrt() + 
  ggtitle("Energy consumption distribution for customer id: 10017936")
```

```{r bothcust, fig.cap = "Quantile plots of energy consumption across hours of the day faceted by weekday/weekend for household 1 (customer id: 10006704) and household 2 (customer id: 10017936). Median behavior for the early morning hours of household 2 is extremely high for weekends compared to weekdays, not much differenc can be observed for household 1. "}
cust1
cust2
```

```{r bothcust-lv, fig.cap ="Violin plots of energy consumption across hours of the day faceted by weekday/weekend for household 2 (customer id: 10017936). Early morning hours are bimodal implying both high and low consumption are probable. Volatility is high from 7 to 13 hours as evident from the narrow violins."}
# 
# smart_meter10 %>% 
#   filter(customer_id %in% c(10006704)) %>% 
#   prob_plot("wknd_wday",
#                             "hour_day",
#                             response = "general_supply_kwh",
#                             plot_type = "violin") +
#   scale_y_sqrt() +
#   ggtitle("Energy consumption distribution for customer id: 10006704")

smart_meter10 %>% 
  filter(customer_id %in% c(10017936))%>% 
  prob_plot("wknd_wday",
                            "hour_day",
                            response = "general_supply_kwh",
                            plot_type = "violin") +
  scale_y_sqrt() + ggtitle("Energy consumption distribution for customer id: 10017936")
```

```{r cust-ridge, fig.cap ="Boxplot of energy consumption across hours of the day faceted by weekday/weekend for household 2 (customer id: 10017936). Median behavior for the early morning hours of household 2 is extremely high for weekends compared to weekdays."}
# 
# smart_meter10 %>% 
#   filter(customer_id %in% c(10006704)) %>% 
#   prob_plot("wknd_wday",
#                             "hour_day",
#                             response = "general_supply_kwh",
#                             plot_type = "violin") +
#   scale_y_sqrt() +
#   ggtitle("Energy consumption distribution for customer id: 10006704")

smart_meter10 %>% 
  filter(customer_id %in% c(10017936))%>% 
  prob_plot("hour_day", 
            "wknd_wday",
            response = "general_supply_kwh",
            plot_type = "boxplot") +
            scale_y_sqrt() +
            # scale_fill_brewer(palette = "Dark2")
  ggtitle("Energy consumption distribution for customer id: 10017936")
```

@@

This case study shows systematic exploration of energy behavior for two random households to gain some insights on periodic behavior of the households. First, it helps us to find the list of cyclic granularities to look at, then shrinks the number of possible visualizations by identifying harmonies, visualise a harmony pair and shows the effect of different distribution plots or reverse mapping.

>>>>

<!-- # how data looks -->
Smart meters provide large quantities of measurements on energy usage for households across Australia. One of the customer trial [@smart-meter] conducted as part of the Smart Grid Smart City (SGSC) project (2010-2014) in Newcastle, New South Wales and some parts of Sydney provides customer wise data on energy consumption for every half hour from February 2012 to March 2014. It would be interesting to explore the energy consumption distribution for these customers and gain insights on their energy behavior which are lost either due to aggregation or looking only at coarser temporal units. The idea here is to show how looking at the energy consumption across different cyclic granularities together can help identify different behavioral patterns.
<!-- and identify the extreme and regular households. -->

@@

The data structure for a sample of 50 households from that trial is as follows:
<!-- We take a random sample of 50 households to see how we can achieve that.  -->

```{r smart-meter, fig.cap = "Smart meter data structure"}
sm <- 
  read_rds("data/sm_cust50.rds")

smart_meter <- sm %>% as_tsibble() %>% dplyr::select(customer_id, 
                              reading_datetime, 
                              general_supply_kwh)
smart_meter
```

@@

@@

```{r search, echo=TRUE}
library(gravitas)
smart_meter %>% search_gran()
```

<!-- #smart_smart_search <- smart_meter %>% search_gran() -->
<!-- #knitr::kable(smart_smart_search, format = "markdown") -->

@@

```{r search_limit1, echo=TRUE}
smart_meter10 %>% 
  search_gran(highest_unit = "month")
```

<!-- # smart_meter_limit1 <- smart_meter10 %>%  -->
<!-- #   search_gran(highest_unit = "month") -->
<!-- # knitr::kable(smart_meter_limit1, format = "markdown") -->

@@

```{r search_gran_limit2, echo=TRUE}
smart_meter10 %>% search_gran(highest_unit = "month",
                              filter_out = c("hhour", "fortnight")
                              )
```

<!-- # smart_meter_limit2 <- smart_meter10 %>% search_gran(highest_unit = "month", -->
<!-- #                               filter_out = c("hhour", "fortnight") -->
<!-- # ) -->
<!-- # knitr::kable(smart_meter_limit2, format = "markdown") -->

@@

<!-- From the Gestalt theory with the same data  <$C1$ $C2$, `general_supply_kwh`>, mapping $C1$ on the x-axis vs facet might lead to different insights. -->
s

```{r harmony, echo=TRUE}
smart_meter10 %>% harmony(
  ugran = "month",
  filter_out = c("hhour", "fortnight")
)
```

<!-- # smart_meter10 %>% gran_advice("wknd_wday", "hour_day") -->

@@

<!-- No need to share the same for 50 customers - cut down text and anyway nothing much interesting is coming -->

<!-- ```{r wknd-wday50, fig.cap = "Quantile plot of energy consumption of 50 households across different hours of the day faceted by weekday and weekend. The quantiles are not different for weekdays and weekends implying either the behavior balances out among these customers or most of them do not behave differently for weekdays and weekends."} -->

<!-- smart_meter <-  -->
<!--   read_rds("data/sm_cust50.rds") -->

<!-- smart_meter %>%  -->
<!--   prob_plot("wknd_wday", -->
<!--                             "hour_day", -->
<!--                             response = "general_supply_kwh", -->
<!--                             plot_type = "quantile", -->
<!--             symmetric = FALSE,  -->
<!--             quantile_prob = c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99)) + -->
<!--   scale_y_sqrt() -->
<!-- ``` -->

@@

<!-- ```{r my-hd, fig.cap="Area quantile plots of energy consumption across hours of the day faceted by months of the year. The black line is the median, whereas the pink band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile. The median and quartile consumption increase in winter months, but extreme behavior represented by higher bands mostly stay the same across seasons. "} -->

<!-- library(gravitas) -->
<!-- library(tsibble) -->

<!-- library(ggplot2) -->
<!-- smart_meter %>% -->
<!--   prob_plot("month_year", "hour_day", -->
<!--     response = "general_supply_kwh", -->
<!--     plot_type = "quantile", -->
<!--     quantile_prob = c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99), -->
<!--     symmetric = TRUE) + -->
<!--   ggtitle("") + scale_y_sqrt(breaks = c(0.01, 0.1, 0.5, 1:3)) -->
<!-- ``` -->

@@

@@

<!-- Boxplot of energy consumption is shown across `wknd_wday` (facet) and `hour_day` (x-axis) for the same two households. For the first household, the shape of the distribution does not change significantly for weekdays and weekends. The distribution is roughly smooth with a morning and an evening peak. -->

In \autoref{fig:bothcust}, the distribution of energy consumption is plotted across the harmony pair (wknd_wday, hour_day) through an area quantile plot. 
The black line is the median, whereas the pink band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile. The first facet represents the weekday behavior while the second one displays the weekend behavior and energy consumption across each hours of the day is shown inside each facet. The energy consumption is extremely skewed with 1st, 10th and 25th percentile lying very close whereas 75th, 90th and 99th lying further away from each other.This is common across both weekdays and weekends. For the first few hours on weekdays, median energy consumption starts and continues to be higher for longer as compared to weekends.

Consider looking at letter value plots instead of quantile plots to look at the same data in \autoref{fig:bothcust-lv}. There is additional information that we can derive looking at the distribution.
There is bimodality in the early hours of the day, implying both low and high energy consumption is probable in the early hours of the day both for weekdays and weekends. Also hours from 7 to 13 look most volatile. If we visualize the same data with reverse mapping of the cyclic granularities, then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. For example in \autoref{fig:cust-lv}, it can be seen that median energy consumption for the early morning hours is extremely high for weekdays compared to weekends. Also, outliers are more prominent in the latter part of the day. All of these indicates that looking at different distribution plots or changing the mapping might shed lights on different aspect of the energy behavior for the same customer.

@@

```{r granadvice, echo = TRUE}
smart_meter10 %>% 
  filter(customer_id %in% c(10017936))%>% 
  gran_advice("wknd_wday",
                            "hour_day")
```

```{r bothcust, fig.cap = "Area quantile plot of energy consumption across hours of the day faceted by weekday/weekend for customer id: 10017936 displays the smoothed energy consumption curve for the entire day. The customer typically consumes more in the morning hours following which their consumption is low till late afternoon hours. Median consumption for the early morning hours starts higher and continue to remain high for longer during weekdays compared to weekdend."}

# plot for only customer is shown
# cust1 <- smart_meter10 %>% 
#filter(customer_id %in% c(10006704)) %>% 
#   prob_plot("wknd_wday",
#                             "hour_day",
#                             response = "general_supply_kwh",
#                             plot_type = "quantile",
#             symmetric = FALSE,
#             quantile_prob = c(0.25, 0.5, 0.75, 0.9)) 
#   scale_y_sqrt() +
#   ggtitle("Energy consumption distribution for customer id: 10006704")
cust2 <- smart_meter10 %>% 
  filter(customer_id %in% c(10017936))%>% 
  prob_plot("wknd_wday",
                            "hour_day",
                            response = "general_supply_kwh",
                            plot_type = "quantile",
            symmetric = TRUE,
             quantile_prob = c(0.01,0.1, 0.25, 0.5, 0.75, 0.9, 0.99)) +
  scale_y_sqrt() 
cust2
```

```{r bothcust-lv, fig.cap ="Violin plots of energy consumption across hours of the day faceted by weekday/weekend for customer id: 10017936. Early morning hours are bimodal implying both high and low consumptions are probable. Bimodality can be caused by difference in behavior across different seasons. Volatility is highest from 7 to 13 hours as evident from the narrow violins."}
# 
# smart_meter10 %>% 
#   filter(customer_id %in% c(10006704)) %>% 
#   prob_plot("wknd_wday",
#                             "hour_day",
#                             response = "general_supply_kwh",
#                             plot_type = "violin") +
#   scale_y_sqrt() +
#   ggtitle("Energy consumption distribution for customer id: 10006704")

smart_meter10 %>% 
  filter(customer_id %in% c(10017936))%>% 
  prob_plot("wknd_wday",
                            "hour_day",
                            response = "general_supply_kwh",
                            plot_type = "violin") +
  scale_y_sqrt() 
```

```{r cust-ridge, fig.cap ="Boxplot of energy consumption across hours of the day faceted by weekday/weekend for customer id: 10017936. The change in mapping of the cyclic granularities leads to easier comparison within weekdays and weekends for each hour. Median consumption till 6am is higher for weekends compared to weekdays although the quartiles look similar. Outliers appear more during the end of the day implying more uncertain behavior in latter hours of the day compared to early morning hours."}
# 
# smart_meter10 %>% 
#   filter(customer_id %in% c(10006704)) %>% 
#   prob_plot("wknd_wday",
#                             "hour_day",
#                             response = "general_supply_kwh",
#                             plot_type = "violin") +
#   scale_y_sqrt() +
#   ggtitle("Energy consumption distribution for customer id: 10006704")

smart_meter10 %>% 
  filter(customer_id %in% c(10017936))%>% 
  prob_plot("hour_day", 
            "wknd_wday",
            response = "general_supply_kwh",
            plot_type = "boxplot") +
            scale_y_sqrt()
            # scale_fill_brewer(palette = "Dark2")

@@

If the data for all keys are visualized together, it might lead to Simpson's paradox, which occurs when one `customer_id` shows a particular behavior, but this behavior is reversed when all keys are combined together. This is also intuitive because `customer_id`'s with very different occupation or demographics will tend to have very different energy behavior and combining them together will somehow weaken any typical or extreme behavior. A strategy for analyzing multiple keys together could be to first screen some basis time series or demographic features and then look at their energy behavior. This is beyond the scope of the current work.

<!-- For the purpose of this paper, the smart meter data is filtered for two customers to illustrate kinds of insights that can be drawn after proceeding with the systematic exploration. -->

<!-- For example, Figure \ref{fig:wknd-wday50} shows the quantile plot of energy consumption of a sample of 50 households. Quantiles for weekends and weekdays for these customers look exactly the same. It is not clear if the behaviors of most of these customers do not alter between weekdays and weekends or looking at all customers together balances out the behaviors. Consider another example in Figure \ref{fig:my-hd} showing the area quantile plots of 50 customers. The median is very close to the lower boundaries of all the other bands implying energy consumption for these households are left skewed. The pink band changes significantly in winter months (May - August) implying that much of the behavioral changes occur in the quartiles, that too in peak hours of the day. It is to be noted here that the the level of all quantiles increased too in winter months (increased energy usage in weather conditions), the interesting part is to notice that the relationship between bands other than quartile stayed same across seasons. -->

This case study shows systematic exploration of energy behavior for a random household to gain some insights on periodic behavior of the households. First, it helps us to find the list of cyclic granularities to look at, then shrinks the number of possible visualizations by identifying harmonies, visualise a harmony pair and shows the effect of different distribution plots or reverse mapping.
